---
title: "10qart"
format: pdf
---

```{r}
candy_file <- url("https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv")

candy = read.csv(candy_file, row.names=1)
head(candy)
```

Q1. How many different candy types?: 
85 different candies, 9 different types?

Q2. How many fruity candy types are in the dataset?
38 fruity type candies

```{r}
sum(candy[, "fruity"])
```
```{r}
## Winpercent of Twix
candy["Twix", ]$winpercent
```

Q3. Favorite candy and winpercent?
It's Kit Kat.

Q4. Winpercent for Kit Kat?

```{r}
candy["Kit Kat", ]$winpercent
```
Q5. Winpercent for Tootsie Rolls?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library("skimr")
skim(candy)

```
Q6. Any variable/column that is on a different scale to the other data?
While most of the variables are in a scale of 0 to 1, "winpercent" is the
only variable with a scale from 0 to 100.


Q7. What does the O and 1 represent in the chocolate column?
0 means the candy does not contain chocolate.
1 means the candy contains chocolate.

Q8. Histogram of winpercent

```{r}
hist(candy[, "winpercent"])
```
Q9. Is the distribution symmetrical?
No. It is slightly skewed to the right.

Q10. Is the center of distribution above or below 50?
It is below 50 as it is right skewed.

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])

mean(candy$winpercent[as.logical(candy$fruity)])
```

Q11. On average is chocolate higher or lower ranked than fruit candy?

Chocolate is ranked higher on average.

Q12. Is this difference statistically significant?

With a p-value below 0.05, this is statistically significant.

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)])
```
```{r}
tail(candy[order(candy$winpercent),], n=5)
head(candy[order(candy$winpercent),], n=5)
```
Q13. Five least liked candy types?
Nik L Nip, Boston Baked Beans, Chiclets, Supper Bubble, Jawbusters

Q14. Top five candies?
Snickers, Kit Kat, Twix, Reese's Mini, Reese's Peanut Butter Cups

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

```{r}
##setting up color vectors for the plot
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

Q17. Worst ranked chocolate candy?
Sixlets

Q18. Best ranked fruity candy?
Starburst


```{r}
library(ggrepel)

##price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head(candy[ord, c(11, 12)], n=5)

tail(candy[ord, c(11, 12)], n=5)
```
Q19. Which candy is highest rank in terms of winpercent for the least money?

Reese's Miniatures are the bang for your buck.

Q20. Top 5 most expensive candies and which is least popular?

Nik L Nip, Nestle Smarties, Ring Pop, Hershey's Krackel, Hershey's Milk Choc

Nik L Nip are the least popular.

```{r}
##first plot
ggplot(candy) + 
  aes(pricepercent, rownames(candy)) +
  geom_col()
```
```{r}
#reordering in descending order
ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy),pricepercent)) +
  geom_col()
```
```{r}
##col to segment + dot lollipop plot
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```
```{r}
library(corrplot)
```
```{r}
cij <- cor(candy)
corrplot(cij)
```
Q22. What two variables are anti-correlated?
Chocolate and Fruit [Weird, I love chocolate + raspberry]

Q23. Which two variables are most positively correlated?
Chocolate and Bar. Great form for chocolate to come in.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)

```
```{r}
##plotting pca score plot pc1 v pc2
plot(pca$x[,1:2])
```

```{r}
##adding color to plot
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
```{r}
##relabling plot with nonoverlapping names
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
##making it interactive
##library(plotly)

##ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
Q24. What original variables are picked up strongly by PC1 in the positive direction? Does this make sense?

Variables picked up in the positive direction are fruity, hard, and pluribus. These are common variables for fruity type candy, so yes this does make sense. 
